Overlapping Batches for the Assessment of Solution Quality in Stochastic Programs

### 2014-06-08 omrp_lb_rev.tex
More Referee 2 and AE comments
minor comments

% Definitions of variables
\newcommand{\X}{X}
\newcommand{\x}{x}
\newcommand{\xh}{{\hat{\x}}}
\newcommand{\xs}{\x^*}
\newcommand{\xit}{\xi}
\newcommand{\xiti}{\xit^i}
\newcommand{\zs}{z^*}
\newcommand{\nb}{\left\lfloor\tfrac{n-m}{\gamma}\right\rfloor+1}
\newcommand{\nbl}{\left\lfloor\tfrac{n_l-m_l}{\gamma_l}\right\rfloor+1}
\newcommand{\gammab}{\bar{\gamma}}
\newcommand{\ogb}{\tfrac{1}{\gammab}}
\newcommand{\cogb}{\left\lceil\ogb\right\rceil}

@@

The overlapping batch means (OBM) method, proposed by \citet{Meketon1984}, is commonly used in simulation to obtain variance estimators having variances that are comparatively lower than their nonoverlapped counterparts. 
In the context of steady-state simulation, OBM is often used to estimate the variance of the sample mean, which is itself an estimator for the mean. 
\cite{SAH90} use overlapping estimators to estimate the variance of estimators for non-means. 
More recently, overlapping estimators based on standardized time series (as opposed to nonoverlapping batch means) have been evaluated \citep{Alexopoulos01012007,Alexopoulos2007}; see also \citep{Meterelliyoz_etal_12}.
To the best of our knowledge, OBM has never before been studied in a stochastic optimization context. 
In this paper we show that the same benefits can be obtained from overlapping batches when used in a stochastic optimization setting.

A natural setting where batch means are used in stochastic optimization (especially in stochastic programming) is for assessing solution quality. 
This typically involves generating point and interval estimators of the optimality gap of a given candidate solution using Monte Carlo simulation. 
Because optimality gap estimators can be difficult to analyze statistically, a common approach is to use nonovelapping batches of data to generate a number of optimality gap estimators. 
These individual estimators are then used to form point and interval estimators applying the classical nonoverlapping batch means method.
This is referred to as the Multiple Replications Procedure (MRP) and was initially discussed in \citep{Mak1999}.   
In this paper, we formulate an ovelapping batches version of MRP, allowing for partial overlap, and provide conditions under which the resulting gap estimators exhibit reduced variance when compared to their counterparts obtained via nonoverlapping batches.
We also show that the resulting point estimators are strongly consistent and the interval estimators are asymptotically valid. 
We present computational results to examine small-sample behavior. 
Our experiments indicate that variance reduction is achieved with small sample sizes, while bias and coverage probability are unaffected.

We consider a stochastic optimization problem of the form 
\begin{align} \tag{SP} \label{eq:sto_prog} 
	z^* & = \min_{\x \in X} \e{f(\x,\xit)},
\end{align}
where $\x$ is a vector of decision variables, $\xit$ is a vector of random variables, and $X$ is the feasible set, which is assumed to be independent of $\xit$.
Further, it is assumed that the distribution of $\xit$ is known and that we can sample from it.
Even though we will impose a more restrictive moment condition later, we assume that $\e{f(\x,\xit)}$ is well defined and finite for all $\x \in X$ and (\ref{eq:sto_prog}) has a finite optimal solution achieved on $X$.

Unless $f$ has a simple structure or the number of realizations of $\xi$ is small, (\ref{eq:sto_prog}) typically cannot be solved exactly. 
An approximation of (\ref{eq:sto_prog}) can be generated by sampling from the distribution of $\xi$.
While other sampling schemes are possible, let $\xit^1, \xit^2, \dots, \xit^m$ be an independent and identically distributed (i.i.d.) sample from the distribution of $\xit$.
A sampling approximation of (\ref{eq:sto_prog}), often called the Sample Average Approximation (SAA), is given by
\begin{align} \tag{SP$_m$} \label{eq:sto_prog_m}
	z_m^* & = \min_{\x \in X} \frac{1}{m} \sum_{i=1}^m f(\x,\xiti).
\end{align}
We denote an optimal solution to (\ref{eq:sto_prog}) as $\xs$ and an optimal solution to (\ref{eq:sto_prog_m}) as $\xs_m$.

We wish to assess the quality of a candidate solution $\xh \in X$. 
Assessing solution quality is important in practice because as (\ref{eq:sto_prog}) typically cannot be solved exactly, one only has an approximate solution $\xh$ without verification of its quality.
Assessing solution quality is also a critical component of stopping criteria in algorithms designed to (approximately) solve (\ref{eq:sto_prog}).
We define the quality of $\xh$ by its optimality gap, $\e{f(\xh,\xit)} - \zs$. 
The lower the optimality gap, the higher the quality of the solution and a zero optimality gap implies that the solution is optimal. 
Note that the candidate solution $\xh \in X$ may have been found in any way. 
We assume that this solution is given as input to our procedures, and hence fixed, throughout this paper.

The optimality gap $\e{f(\xh,\xit)} - \zs$ often cannot be calculated exactly. First, for a given candidate solution $\xh \in X$, evaluation of $\e{f(\xh,\xit)}$ typically involves a difficult multidimensional integral. 
Second, we do not know $\zs$. 
In many fields of optimization, the second problem is alleviated by obtaining lower bounds on $\zs$ through relaxations such as integrality, Lagrangian or semidefinite relations.
As a result, it is common to evaluate an upper bound on the optimality gap. 
This motivates estimating the upper bound on the optimality gap in stochastic programming by using Monte Carlo simulation.
Given a sample size $m$, an upper bound on the optimality gap can be obtained by $$
\e{f(\xh,\xit)} - \zs \leq \e{f(\xh,\xit)} - \e{\zs_m}
$$ 
due to the inequality $\e{\zs_m} \leq \zs$ \citep{Mak1999,norkin_pflug_ruszczynski_98}.
This bound improves as the sample size increases, that is, $\e{\zs_m} \leq \e{\zs_{m+1}} \leq \zs$.  
A straightforward estimate of $\e{f(\xh,\xit)}$ is the sample mean, $\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti)$. 
Instead of $\e{\zs_m}$, simply $\zs_m$ can be used.  
The resulting point estimator of optimality gap is then 
$$
\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti) - \zs_m.
$$ 
Here, we assume the same observations $\xit^1, \dots, \xit^m$ are used in both terms.

Computing the above optimality gap estimate involves solving an optimization problem (\ref{eq:sto_prog_m}) to obtain a lower bound estimator of $\zs$, $\zs_m$, which complicates the statistical analysis. 
As mentioned above, to enable statistical inference, the MRP of \citet{Mak1999} generates $k$ independent estimators, each using sample size $m$ ($k$ nonoverlapping batches of size $m$) and averages them to obtain a point estimator.  
To form confidence intervals, the sample variance of these estimators is calculated (we further review MRP in \S \ref{ssec:mrp}.)  
An advantage of MRP is its applicability to a wide range of problems.  
With i.i.d.\ sampling, $f$ can be linear or nonlinear, $\X$ can include integrality constraints or not.  
It is also easy to implement; thus, MRP has been applied to a wide variety of problems, including finance, supply chain network design, etc.; see e.g., \citep{bertocchi_etal_99,janjarassuk_linderoth_08,santoso_ahmed_etal_05}.  
Recently, the approach of (nonoverlapping) batching has been used for assessing solution quality of stochastic programs with finitely many expected value \citep{wang_ahmed_08} and stochastic dominance \citep{hu2010sample} constraints.

>>>>

% Definitions of variables
\newcommand{\X}{X}
\newcommand{\x}{\mathbf{x}}
\newcommand{\xh}{{\hat{\x}}}
\newcommand{\xs}{\x^*}
\newcommand{\xit}{\boldsymbol{\xi}}
\newcommand{\xiti}{\xit^i}
\newcommand{\zs}{z^*}
\newcommand{\nb}{\left\lfloor\tfrac{n-m}{\gamma}\right\rfloor+1}
\newcommand{\nbl}{\left\lfloor\tfrac{n_l-m_l}{\gamma_l}\right\rfloor+1}
\newcommand{\gammab}{\bar{\gamma}}
\newcommand{\ogb}{\tfrac{1}{\gammab}}
\newcommand{\cogb}{\left\lceil\ogb\right\rceil}

@@

The Overlapping Batch Means (OBM) method, proposed by \citet{Meketon1984}, is commonly used in simulation to obtain variance estimators having variances that are comparatively lower than their nonoverlapped counterparts. 
In the context of steady-state simulation, OBM is often used to estimate the variance of the sample mean, which is itself an estimator for the mean. 
\cite{SAH90} use overlapping estimators to estimate the variance of estimators for non-means. 
More recently, overlapping estimators based on standardized time series (as opposed to batch means) have been evaluated \citep{Alexopoulos01012007,Alexopoulos2007}; see also \citep{Meterelliyoz_etal_12}.
To the best of our knowledge, OBM has never before been studied in a stochastic optimization context. 
In this paper we show that the same benefits can be obtained from overlapping batches when used in a stochastic optimization setting.

A natural setting where batch means are used in stochastic optimization (especially in stochastic programming) is for assessing solution quality. 
This typically involves generating point and interval estimators of the optimality gap of a given candidate solution using Monte Carlo simulation. 
Because optimality gap estimators can be difficult to analyze statistically, a common approach is to use nonovelapping batches of data to generate a number of optimality gap estimators. 
These individual estimators are then used to form point and interval estimators applying the classical nonoverlapping batch means method.
This is referred to as the Multiple Replications Procedure (MRP) and was initially discussed in \citep{Mak1999}.   
In this paper, we formulate an ovelapping batches version of MRP, allowing for partial overlap, and provide conditions under which the resulting gap estimators exhibit reduced variance when compared to their counterparts obtained via nonoverlapping batches.
We also show that the resulting point estimators are strongly consistent and the interval estimators are asymptotically valid. 
We also present computational results to examine small-sample behavior. 
Our experiments indicate that the use of overlapping estimators results in gap estimators achieving lower variance than their nonoverlapping counterparts, along with comparable bias and interval coverage---for both small- and large-sample cases.

We consider a stochastic optimization problem of the form 
\begin{align} \tag{SP} \label{eq:sto_prog} 
	z^* & = \min_{\x \in X} \e{f(\x,\xit)},
\end{align}
where $\x$ is a vector of decision variables, $\xit$ is a vector of random variables, and $X$ is the feasible set, which consists of only deterministic constraints. 
There are no stochastic constraints such as probabilistic or expected-value constraints. 
Further, it is assumed that the distribution of $\xit$ is known and that we can sample from it.
Even though we will impose a more-restrictive moment condition later, we assume that $\e{f(\x,\xit)}$ is well defined and finite for all $\x \in X$ and (\ref{eq:sto_prog}) has a finite optimal solution achieved on $X$.

Unless $f$ has a simple structure or the number of realizations of $\xi$ is small, (\ref{eq:sto_prog}) typically cannot be solved exactly. 
An approximation of (\ref{eq:sto_prog}) can be generated by sampling from the distribution of $\xit$.
While other sampling schemes are possible, let $\xit^1, \xit^2, \dots, \xit^m$ be an independent and identically distributed (i.i.d.) sample from the distribution of $\xit$.
A sampling approximation of (\ref{eq:sto_prog}), often called the sample average approximation, is given by
\begin{align} \tag{SP$_m$} \label{eq:sto_prog_m}
	z_m^* & = \min_{\x \in X} \frac{1}{m} \sum_{i=1}^m f(\x,\xiti).
\end{align}
We denote an optimal solution to (\ref{eq:sto_prog}) as $\xs$ and an optimal solution to (\ref{eq:sto_prog_m}) as $\xs_m$.

We wish to assess the quality of a candidate solution $\xh \in X$. 
Assessing solution quality is important in practice because as (\ref{eq:sto_prog}) typically cannot be solved exactly, one only has an approximate solution $\xh$ without verification of its quality.
Assessing solution quality is also a critical component of stopping criteria in algorithms designed to (approximately) solve (\ref{eq:sto_prog}).
We define the quality of $\xh$ by its optimality gap, $\e{f(\xh,\xit)} - \zs$. 
The lower the optimality gap, the higher the quality of the solution; and a zero optimality gap implies that the solution is optimal. 
Note that the candidate solution $\xh \in X$ may have been found in any way. 
We assume that this solution is given as input to our procedures, and hence fixed, throughout this paper.

The optimality gap $\e{f(\xh,\xit)} - \zs$ often cannot be calculated exactly. First, for a given candidate solution $\xh \in X$, evaluation of $\e{f(\xh,\xit)}$ typically involves a difficult multidimensional integral. 
Second, we do not know $\zs$. 
In many fields of optimization, the second problem is alleviated by obtaining lower bounds on $\zs$ through relaxations such as integrality, Lagrangian, or semidefinite relations.
As a result, it is common to evaluate an upper bound on the optimality gap. 
This motivates estimating the upper bound on the optimality gap in stochastic programming by using Monte Carlo simulation.
Given a sample size $m$, an upper bound on the optimality gap can be obtained by $$
\e{f(\xh,\xit)} - \zs \leq \e{f(\xh,\xit)} - \e{\zs_m}
$$ 
due to the inequality $\e{\zs_m} \leq \zs$ \citep{Mak1999,norkin_pflug_ruszczynski_98}.
This bound improves as the sample size increases, that is, $\e{\zs_m} \leq \e{\zs_{m+1}} \leq \zs$.  
A straightforward estimate of $\e{f(\xh,\xit)}$ is the sample mean, $\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti)$. 
Instead of $\e{\zs_m}$, we can simply use $\zs_m$.  
The resulting point estimator of the optimality gap is then 
$$
\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti) - \zs_m.
$$ 
Here, we assume the same observations $\xit^1, \dots, \xit^m$ are used in both terms.

Computing the above optimality gap estimate involves solving an optimization problem (\ref{eq:sto_prog_m}) to obtain a lower bound estimator of $\zs$, $\zs_m$, which complicates the statistical analysis. 
As mentioned above, to enable statistical inference, the MRP of \citet{Mak1999} generates $k$ independent estimators, each using sample size $m$ ($k$ nonoverlapping batches of size $m$) and averages them to obtain a point estimator.  
The sample variance of these estimators is used to form confidence intervals for the optimality gap (we further review MRP in \S \ref{ssec:mrp}).  
An advantage of MRP is its applicability to a wide range of problems.  
With i.i.d.\ sampling, $f$ can be linear or nonlinear, $\X$ can include integrality constraints or not.  
It is also easy to implement; thus, MRP has been applied to a wide variety of problems, including finance, supply chain network design, etc.; see e.g., \citep{bertocchi_etal_99,janjarassuk_linderoth_08,santoso_ahmed_etal_05}.  
Recently, the approach of (nonoverlapping) batching has been used for assessing solution quality of stochastic programs with finitely many expected value \citep{wang_ahmed_08} and stochastic dominance \citep{hu2010sample} constraints.
